---
title: 'Basic Anova Examples'
output: pdf_document
---

#Azevedo C.V.G - cvg.azevedo@gmail.com

###-----ANOVA ASSUMPTIONS-----###

# 1) Normality of errors (Errors must follow Normal distribution)
# 2) Homoscedasticity of errors (Errors must be homogeneous)
# 3) Independence from errors (Errors must be independent)
# 4) Additivity of the model (The model must contain only additive effects)

## Packages
```{r}
library("agricolae")
library("tidyverse")
library("laercio")
library("fBasics")
library("car")
library('asbio')
library('MASS')
library("outliers")
library('emmeans')
library('knitr')
```

## SET WORKING DIRECTORY
```{r}
rm(list = ls())
setwd('C:\\Users\\clebe\\Dropbox\\TOSHIBA EXT.dbx-external-drive\\Cleber\\Google Drive (not syncing)\\R\\Course\\leituraDados')
```

## DATASET 
```{r}
data(cotton)
rdata<-cotton # change the name to your dataset
head(rdata)
str(rdata)
```

## Frequency Histogram
```{r}
bw = 1 #binwidth to normal curve
ggplot(rdata, aes(x=yield))+
  geom_histogram(aes(y=..density..),bins=12, colour='black',fill='lightgray')+
  geom_density(aes(y=..density..), colour='blue')+
  stat_function(fun = function(x) 
    dnorm(x, mean = mean(rdata$yield), sd = sd(rdata$yield)), colour='red' )+
    theme_classic()+
labs(x='Yield', y='Frequency') #Label names
```
## CHECKING OUTLIERS
```{r}
boxplot(rdata$yield)

outlier(
  x = rdata$yield,
  logical = TRUE
) %>% as_tibble %>% 
  bind_cols(rdata)
```

## CHECKING ASSUMPTIONS
# Extracting Residuals - DIAGNOSTIC ANOVA
```{r}
rdata$lineage <- as.factor(rdata$lineage)
D.AOV<-lm(yield ~ lineage + block, data=rdata) # FIT YOUR MODEL
#summary(D.AOV)
#D.AOV$residuals
```

# Checking NORMALITY - Shapiro-Wilk
```{r}
shapiro.test(D.AOV$residuals)
qqnormPlot(D.AOV$residuals)
```

# Checking Homoscedasticity - Barlett
```{r}
bartlett.test(D.AOV$residuals~lineage,data=rdata) 
```
# Checking INDEPENDENCE - Durbin-Watson Test
```{r}
dwt(lm(yield ~ lineage + block, data=rdata))
```
# Checking Additivity
```{r}
tukey.add.test(rdata$yield, rdata$lineage, rdata$block)
```
## IF MEETING THE ASSUMPTIONS - ANOVA
```{r}
anova(D.AOV)
```
## IF NOT MEETING ASSUMPTIONS - DATA TRANSFORMATION 
# Testing Common Transformations
```{r}
#Square root
rdata$SQRT<-rdata$yield^(1/2)

#Cubic root
rdata$CUB<-rdata$yield^(1/3)

#Log
rdata$LOG<-log(rdata$yield)

#Power
rdata$POT2<-rdata$yield^2

#New data frame
head(rdata)
```

## Checking Assumptions
```{r}
AOVSQRT<-lm(SQRT ~ lineage, data=rdata)
AOVCUB<-lm(CUB ~ lineage, data=rdata)
AOVLOG<-lm(LOG ~ lineage, data=rdata)
AOVPOT2<-lm(POT2 ~ lineage, data=rdata)

#Square Root
shapiro.test(AOVSQRT$residuals)
qqnormPlot(AOVSQRT$residuals)
bartlett.test(residuals(AOVSQRT)~rdata$lineage)
car::dwt(lm(SQRT ~ lineage, data=rdata))

#Cubic Root
shapiro.test(AOVCUB$residuals)
qqnormPlot(AOVCUB$residuals)
bartlett.test(residuals(AOVCUB)~rdata$lineage)
car::dwt(lm(CUB ~ lineage, data=rdata))

#LOG
shapiro.test(AOVLOG$residuals)
qqnormPlot(AOVLOG$residuals)
bartlett.test(residuals(AOVLOG)~rdata$lineage)
car::dwt(lm(LOG ~ lineage, data=rdata))

#POWER2
shapiro.test(AOVPOT2$residuals)
qqnormPlot(AOVPOT2$residuals)
bartlett.test(residuals(AOVPOT2)~rdata$lineage)
car::dwt(lm(POT2 ~ lineage, data=rdata))
```

## Transformation Through BOX-COX
```{r}
bc<-boxcox(D.AOV)
bc

lambda <- bc$x[which(bc$y==max(bc$y))]
lambda
```

# Transformation using lambda value
```{r}
rdata$bc<-rdata$yield^lambda
head(rdata)
```

# Checking 
```{r}
AOVbc<- lm(bc ~ lineage + block, data=rdata)
shapiro.test(residuals(AOVbc))
bartlett.test(residuals(AOVbc)~rdata$lineage)
car::dwt(lm(AOVbc)) 
asbio::tukey.add.test(rdata$bc, rdata$block, rdata$lineage)
```
# ANOVA
```{r}
anova <- anova(AOVbc)
cv.model(AOVbc)
print(HSD.test(AOVbc, "lineage", group = TRUE)$groups)
plot(HSD.test(AOVbc, "lineage", group = TRUE))
lsmeans <- lsmeans(AOVbc,'lineage', adjust='t')
kable(anova, digits=3,)
kable(lsmeans, digits=2)
```

#GRAPH
```{r}
ggplot(rdata, aes(x = lineage, y = yield, color=lineage)) +
  geom_boxplot() +
  scale_colour_manual(position=("left"), values=c("blue", "green", "orange","red")) +
    guides(color = FALSE) +
  theme_classic() +
  labs(x="Genotype", y="Yield")
```
#Writing .TXT file 
```{r}
sink("ANOVA.txt",type = c("output"))
cat("#---- ANOVA ----#", "\n")
print(anova(AOVbc))
cat("#--- CV% ---#", "\n")
print(cv.model(AOVbc))
cat("#--- Tukey ---#", "\n")
print(HSD.test(AOVbc, "lineage", group = TRUE)$groups)
sink()
```
#Resume Anova
```{r}
anova(AOVbc)
cat("#--- CV% ---#", "\n")
agricolae::cv.model(AOVbc)
```

#Quantitative example
```{r}
brix<-read.table("https://raw.githubusercontent.com/VSSEric/Introduction-to-R-UFLA-2019/master/Exemplos/dbc2.txt", h=T)
str(brix)
brix<-transform(brix, Bloco=factor(Bloco))
str(brix)
```
#Box Plot
```{r}
boxplot(brix)
```
#Outliers list
```{r}
outlier(
  x = brix$Brix,
  logical = TRUE
) %>% as_tibble %>% 
  bind_cols(brix)
```

#Plot
```{r}
require(lattice)
xyplot(Brix~Epoca, data=brix, jitter.x=T, type=c("p","a"))
```
#Model
```{r}
mbrix<-lm(Brix ~ Bloco + ordered(Epoca), data=brix)
```

#Test assumptions
```{r}
shapiro.test(mbrix$residuals)
bartlett.test(residuals(mbrix)~brix$Epoca)
car::dwt(lm(Brix ~ Bloco + ordered(Epoca), data=brix))
asbio::tukey.add.test(brix$Brix, brix$Bloco, ordered(brix$Epoca))
```
#Anova
```{r}
anova(mbrix)
agricolae::cv.model(mbrix)
```
#Summary
```{r}
summary(mbrix)
```
#Test effects
```{r}
m1brix<-lm(Brix ~ Bloco + Epoca, data=brix)
m2brix<-lm(Brix ~ Bloco + Epoca + I(Epoca^2), data=brix)
m3brix<-lm(Brix ~ Bloco + Epoca + I(Epoca^2) + I(Epoca^3), data=brix)
m4brix<-lm(Brix ~ Bloco + Epoca + I(Epoca^2) + I(Epoca^3) + I(Epoca^4), data=brix)
m5brix<-lm(Brix ~ Bloco + Epoca + I(Epoca^2) + I(Epoca^3) +I(Epoca^4) +I(Epoca^5), data=brix)
```

#Anova
```{r}
anova(m1brix)
anova(m2brix)
anova(m3brix)
anova(m4brix)
anova(m5brix)
```
#Anova effects
```{r}
anova(m1brix, m2brix, m3brix, m4brix, m5brix)
```
#Best model
```{r}
summary(m2brix)
```
#Means
```{r}
Majus<-predict(m2brix)
ajust<-as.data.frame(Majus)
ajust$Epoca<-brix$Epoca
ajust
```

#Model
```{r}
m1<-lm(Majus ~ Epoca + I(Epoca^2), data=ajust)
```

#Anova
```{r}
anova(m1)
```
#Summary
```{r}
summary(m1)
```
#Coefficient
```{r}
coeficientes<-coef(m1)
coeficientes
```
#Plots
```{r}
library("ggplot2")
library("ggpmisc")
ggplot(ajust, aes(x=Epoca, y=Majus)) +
  stat_smooth(data=ajust, aes(y = Majus, x=Epoca),method = "lm", formula = y ~ x + I(x^2), size = 1)+
  stat_poly_eq(aes(label =  paste(..eq.label.., ..rr.label.., sep = "~~~~")),formula = y ~ x + I(x^2), parse = TRUE) +
  geom_point(size=1.5, pch=21, fill='darkred',
             colour='grey30', alpha=0.7) +
  scale_x_continuous(limits=c(0,35), breaks=seq(0,35,1)) +
  scale_y_continuous(limits=c(10,25), breaks=seq(10, 25, 1)) +
  xlab('Época (Dias)') +
  ylab(expression(paste('Brix (' , degree, 'Bx)'))) +
  theme_bw()
```
#Writing file .TXT with results
```{r}
sink("Resultados das Análises - DBC QUANTITATIVO.txt",type = c("output"))

cat("#---- TABELA DE ANÁLISE DE VARIÂNCIA ----#", "\n")
print(anova(mbrix))
cat("#--- CV% ---#", "\n")
print(cv.model(mbrix))
cat("#---- TABELA DE ANÁLISE DE VARIÂNCIA COMPLETO ----#", "\n")
print(anova(m5brix))
cat("#---- ANÁLISE DO MODELO INDICADO ----#", "\n")
print(anova(m1brix, m2brix, m3brix, m4brix, m5brix))
cat("#---- ANOVA DO MODELO INDICADO ----#", "\n")
print(anova(m2brix))
cat("#---- TABELA DE ANÁLISE DE VARIÂNCIA DO MODELO AJUSTADO ----#", "\n")
print(anova(m1))
cat("#--- Coeficientes do modelo quadratico ajustado ---#", "\n")
print(coef(m1))
cat("#--- R2 ---#", "\n")
print(summary(m1)$r.squared)

sink()
```

#Excel 
```{r}
write.table(ajust, file='Médias Ajustadas (Brix).csv', 
            row.names=FALSE, col.names=c("Média(Brix)", "Época"), sep =" ")  
```

#Using ExpDes.pt Package
```{r}
library(ExpDes.pt)
AnovaDBC<-dbc(trat=brix$Epoca, bloco=brix$Bloco, resp=brix$Brix, quali = FALSE)
graficos(AnovaDBC, grau = 2, mod = TRUE, main = " ", sub = " ",
         xlab = "Época", ylab = "Brix", pch = 19,
         xlim = NULL, ylim = NULL, bty = "o")

```
#Qualitative example
```{r}
#Pacote com conjunto de dados
install.packages("agridat")
```
#Package and opening the data
```{r}
library("agridat")
DADOS<-(dasilva.maize)
```

#Wornking with one environment
```{r}
DADOSEXP1<-subset(DADOS, DADOS$env=="E1")
head(DADOSEXP1)
tail(DADOSEXP1)
```
#BoxPlot
```{r}
boxplot(DADOSEXP1$yield)
```
#Getting outliers
```{r}
outlier(
  x = DADOSEXP1$yield,
  logical = TRUE
) %>% as_tibble %>% 
  bind_cols(DADOSEXP1)
```

#Plot
```{r}
require(lattice)
xyplot(yield~gen, data=DADOSEXP1, jitter.x=T, type=c("p","a"))
```
#Anova
```{r}
analise1<-aov(yield ~ rep + gen, data=DADOSEXP1)
```

#Assumptions
```{r}
shapiro.test(analise1$residuals)
bartlett.test(residuals(analise1)~DADOSEXP1$gen)
car::dwt(lm(yield ~ rep + gen, data=DADOSEXP1))
asbio::tukey.add.test(DADOSEXP1$yield, DADOSEXP1$rep, DADOSEXP1$gen)
```
#Anova
```{r}
anova(analise1)
agricolae::cv.model(analise1)
```
#Tukey test and plot
```{r}
print(agricolae::HSD.test(analise1, "gen", group = TRUE)$groups)
plot(agricolae::HSD.test(analise1, "gen", group = TRUE))
```
#Scott-Knott test
```{r}
library("ScottKnott")
agrupamento<-SK(analise1, which="gen")
summary(agrupamento)
plot(agrupamento, title='Agrupamento dos Genótipos de Milho - Ambiente 1')

```
#Plot
```{r}
library("ggplot2")
ggplot(DADOSEXP1, aes(x = gen, y = yield, fill=gen)) +
  geom_boxplot() +
  scale_y_continuous(limits=c(1, 10)) +
  guides(fill = FALSE) +
  labs(title = "Produtividade média de genótipos de milho - Ambiente 1",
       y = expression(Produtividade*" "(Kg*" "*ha^-1)), x = "Genótipos",
       caption = "Fonte: Curso Gen 2020")
```

#Creating a data frame with the results of SK
```{r}
rank<-data.frame(summary(agrupamento))
colnames(rank)<-c("Genotipos", "Media", "Letra")
rank
```

#Color pallet -> 3 grupos
```{r}
cores <- c(rgb(143,199,74,maxColorValue = 255),
           rgb(242,104,34,maxColorValue = 255), 
           rgb(111,145,202,maxColorValue = 255))
```

#Environment 2
```{r}
DADOSEXP2<-subset(DADOS, DADOS$env=="E2")
```

```{r}
outlier(
  x = DADOSEXP2$yield,
  logical = TRUE
) %>% as_tibble %>% 
  bind_cols(DADOSEXP2)
```

```{r}
require(lattice)
xyplot(yield~gen, data=DADOSEXP2, jitter.x=T, type=c("p","a"))
```

```{r}
analise2<-aov(yield ~ rep + gen, data=DADOSEXP2)
```

```{r}
shapiro.test(analise2$residuals)
bartlett.test(residuals(analise2)~DADOSEXP2$gen)
car::dwt(lm(yield ~ rep + gen, data=DADOSEXP2))
asbio::tukey.add.test(DADOSEXP2$yield, DADOSEXP2$rep, DADOSEXP2$gen)
```

```{r}
anova(analise2)
agricolae::cv.model(analise2)
```

```{r}
print(agricolae::HSD.test(analise2, "gen", group = TRUE)$groups)
plot(agricolae::HSD.test(analise2, "gen", group = TRUE))
```

```{r}
library("ScottKnott")
agrupamento2<-SK(analise2, which="gen")
summary(agrupamento2)
plot(agrupamento2, title='Agrupamento dos Genótipos de Milho - Ambiente 2')
```

```{r}
library("ggplot2")
ggplot(DADOSEXP2, aes(x = gen, y = yield, fill=gen)) +
  geom_boxplot() +
  scale_y_continuous(limits=c(1, 10)) +
  guides(fill = FALSE) +
  labs(title = "Produtividade média de genótipos de milho - Ambiente 2",
       y = expression(Produtividade*" "(Kg*" "*ha^-1)), x = "Genótipos",
       caption = "Fonte: Curso Gen 2020")
```


```{r}
rank2<-data.frame(summary(agrupamento2))
colnames(rank2)<-c("Genotipos", "Media", "Letra")
```

#Análise via ExpDes.pt
```{r}
DBCAMB1<-dbc(DADOSEXP1$gen, DADOSEXP1$rep, DADOSEXP1$yield, quali = TRUE, mcomp = "sk")
```

```{r}
DBCAMB2<-dbc(DADOSEXP2$gen, DADOSEXP2$rep, DADOSEXP2$yield, quali = TRUE, mcomp = "sk")
```

#Multi environment analisys
$$
y_{ijk} = \mu + b_{j(k)} + a_k + t_i + ta_{ik} + e_{ijk}
$$

```{r}
DADOS
str(DADOS)
```

```{r}
boxplot(DADOS$yield)
```


```{r}
conjunta<-aov(yield ~ rep%in%env + gen*env, data=DADOS)
```

```{r}
anova(conjunta)
agricolae::cv.model(conjunta)
```

```{r}
car::Anova(lm(yield ~ env:rep + env + gen + gen*env, data=DADOS)) #add type=3
```
#Interaction Plots
```{r}
with(DADOS, interaction.plot(gen, env, yield, ylab = "médias",
   xlab = "Genótiposs"))

with(DADOS, interaction.plot(env, gen, yield, ylab = "médias",
   xlab = "Ambientes"))
```


```{r}
library(emmeans)
mg<-emmeans(conjunta, ~ gen, data=DADOS)
me<-emmeans(conjunta, ~ env, data=DADOS)
mge<-emmeans(conjunta, ~ gen*env, data=DADOS)
```

#Summary
```{r}
m<-data.frame(mge)
m[,4:7]<-NULL
n<-data.frame(mg)
n[,3:6]<-NULL
q<-data.frame(me)
q[,3:6]<-NULL
head(n)
head(q)
head(m)
```

#Genotypes Ranking 
```{r}
ggplot(n, aes(x = gen, y = emmean)) +
  geom_col(aes(fill=-(emmean))) +
  ylim(0,10)+
  guides(fill = FALSE)+
  labs(title = "Produtividade média geral dos genótipos de milho",
       y = expression(Produtividade*" "(Kg*" "*ha^-1)), x = "Genótipos",
       caption = "Fonte: Curso Gen 2020")
```

#Environment Ranking 
```{r}
ggplot(q, aes(x = env, y = emmean)) +
  geom_col(aes(fill=-(emmean))) +
  ylim(0,15)+
  guides(fill = FALSE)+
  labs(title = "Produtividade média geral dos genótipos de milho por ambiente",
       y = expression(Produtividade*" "(Kg*" "*ha^-1)), x = "Ambiente",
       caption = "Fonte: Curso Gen 2020")
```
#Interaction Graph
```{r}
library(reshape2)
m = melt(m, value.name="emmean")
ggplot(data=m, aes(x=env, y=emmean, group = gen, colour = gen)) +
  geom_line() +
  labs(title = "Produtividade média dos genótipos de milho",
       y = expression(Produtividade*" "(Kg*" "*ha^-1)), x = "Ambientes", 
       colour = "Genótipos",
       caption = "Fonte: Curso Gen 2020")
```
#Multi-Environment Traisl Analysis (metan) - Tiago Olivoto (2020)
```{r}
library(metan)
```
```{r}
mod<-anova_joint(DADOS, env, gen, rep, yield)
```
#Adjusted means
```{r}
mediasge<-ge_means(DADOS, env, gen, resp = yield)
```

#Ranking
```{r}
ge_winners(DADOS, env, gen, resp=yield, type = "ranks")
ge_winners(DADOS, env, gen, resp=yield, type = "winners")
```

```{r}
ge_plot(DADOS, env, gen, yield, type = 2)
```